Well-quasi-orderings on word languages

### 2024-08-09 src/conclusion.tex
starting to add amalgamation systems

\subparagraph{Towards infinite alphabets} Can we go to
alphabets that are infinite but equipped with a well-quasi order?

\subparagraph{Monoid equations}

>>>>

We provided concretes statements that justify why the \kl{subword relation}
is used when defining \kl{well-quasi-orders} on finite words. Even if
\kl{prefix}, \kl{suffix} or \kl{infix} relations are meaningful, they are
\kl{well-quasi-ordered} if and only if they behave similarly to disjoint copies
of $\Nat$ or $\Nat^2$. However, our approach suffers some limitations 
and opens the road to natural continuation of this line of work.

\subparagraph{Towards infinite alphabets} In this paper, we restricted our
attention to \emph{finite} alphabets, having in mind the application to regular
languages. However, the conclusions of \cref{infix-finite-automata:thm} and
\cref{prefixes:thm} could be conjecture to hold in the case of infinite
alphabets (themselves equipped with a \kl{well-quasi-ordering}). This would
require new techniques, as the finiteness of the alphabet is crucial to all of
our positive results.

\subparagraph{Monoid equations}  It could be interesting to understand which
monoids $M$ recognize languages that are \kl{well-quasi-ordered} by the
\kl{infix}, \kl{prefix} or \kl{suffix} relations. This is connected to finding
which classes of graphs of \emph{bounded clique-width} are
\kl{well-quasi-ordered} with respect to the \emph{induced subgraph relation}, a
subject that is recently gaining traction.

### 2024-08-09 src/infixes.tex
starting to add amalgamation systems

\begin{theorem}
    \label{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

\begin{itemize}
    \item Define amalgamation systems
    \item Pumping argument
    \item Characterization
    \item Decision procedure
\end{itemize}

>>>>

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

\begin{definition}
    \textbf{TODO}. An \intro{amalgamation system}
    is a triple such that.
\end{definition}

Let us recall some examples of \kl{amalgamation systems}
arising in the literature.

\begin{example}
    Regular languages, context-free languages, and languages recognized by
    VASS are all recognized by amalgamation systems.
\end{example}

\begin{lemma}
    Let $L$ be a language recognized by an \kl{amalgamation system}
    and assume that $L$ is \kl{well-quasi-ordered}
    by the \kl{infix relation}. Then,
    for every minimal run $\rho$,
    the \kl{gap languages} of $\rho$
    are \kl{subsemigroup languages}.
\end{lemma}

\begin{proofof}{infix-amalgamation:thm}
    TODO.
\end{proofof}

\begin{lemma}
    Let $L$ be given by an \kl{effective amalgamation system},
    then it is decidable whether $L$ 
    is \kl{well-quasi-ordered} by the \kl{infix relation}.
\end{lemma}

### 2024-08-09 src/introduction.tex
starting to add amalgamation systems

Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. The three alternatives
we consider are the \kl{prefix relation}, the \kl{suffix relation}, and the
\kl{infix relation}. Unfortunately, it is an easy check that all of these
constructions are not well-quasi-ordered as soon as the alphabet contains two
non-equivalent letters. This is not the end of the road, because some
\emph{subsets} of $\Sigma^*$ might still be \kl{well-quasi-ordered} by these
relations, the simplest example being the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing the subsets $L \subseteq \Sigma^*$
that are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

>>>>

Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$,
written $u \intro*\higleq v$,
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. In our, case we
consider alphabets $\Sigma$ that are finite and equipped with the equality
relation. In this setting, the three alternatives we consider are the
\intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with $uw
= v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there exists
$w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---.
Unfortunately, it is an easy check that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters.
This is not the end of the road, because some \emph{subsets} of $\Sigma^*$
might still be \kl{well-quasi-ordered} by these relations, the simplest example
being the case of finite sets of (finite) words which are all
\kl{well-quasi-ordered} regardless of the ordering considered. This raises the
question of characterizing the subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} with respect to the \kl{prefix relation} (respectively,
the \kl{suffix relation} or the \kl{infix relation}), and providing suitable
decision procedures.

### 2024-08-09 src/prefixes.tex
starting to add amalgamation systems

In this section, we study the well-quasi-ordering of languages under the prefix
relation. A word $u$ is a \intro{prefix} of a word $w$ if there exists a word
$v$ such that $w = uv$. We denote this relation by $u \intro*\prefleq w$.
Similarly, a word $u$ is a \intro{suffix} of a word $w$ if there exists a word
$v$ such that $w = vu$. We denote this relation by $u \intro*\suffleq w$. Let
us immediately remark that the map $u \mapsto u^R$ that reverses a word is an
order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$. Therefore, we
will focus on the prefix relation in the rest of this section.

@@

\begin{corollary}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}
\begin{proof}
    If $L$ is regular, then it is \kl{$\MSO$-definable}, and there 
    exists a formula $\varphi(x)$ in \kl{$\MSO$} that selects nodes 
    of the \kl{tree of prefixes} that belong to $L$. Now, to decide whether there
    exists an \kl{antichain branch} for $L$, we can simply check whether
    the following formula is satisfied:
    \begin{equation*}
        \exists B. 
        B \text{ is a branch } \land
        \forall x \in B, \exists y. y \text{ is a child of } x \land \varphi(y) \land y \not\in B
        \quad .
    \end{equation*}
    Because the above formula is an $\MSO$-formula over the infinite
    binary tree, whether it is satisfied is decidable.
    \textbf{TODO: cite}.
\end{proof}

@@

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be an infinite language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists finite numbers $k,\ell \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $\ell$.
\end{corollary}
\begin{proof}
    \textbf{todo.}
\end{proof}

Let us conclude by noting that it is unsurprisingly not possible to decide
whether a decidable language is \kl{well-quasi-ordered} by the \kl{prefix
relation}.

\begin{lemma}
    todo.
\end{lemma}

>>>>

In this section, we study the well-quasi-ordering of languages under the prefix
relation. Let us immediately remark that the map $u \mapsto u^R$ that reverses
a word is an order-bijection between $(X^*, \prefleq)$ and $(X^*, \suffleq)$.
Therefore, we will focus on the prefix relation in the rest of this section.

@@

\begin{corollary}
    \label{prefix-wqo-reg-decidable:cor}
    If $L$ is regular, then the existence of an infinite antichain is decidable.
\end{corollary}
\begin{proof}
    If $L$ is regular, then it is \kl{$\MSO$-definable}, and there 
    exists a formula $\varphi(x)$ in \kl{$\MSO$} that selects nodes 
    of the \kl{tree of prefixes} that belong to $L$. Now, to decide whether there
    exists an \kl{antichain branch} for $L$, we can simply check whether
    the following formula is satisfied:
    \begin{equation*}
        \exists B. 
        B \text{ is a branch } \land
        \forall x \in B, \exists y. y \text{ is a child of } x \land \varphi(y) \land y \not\in B
        \quad .
    \end{equation*}
    Because the above formula is an $\MSO$-formula over the infinite
    binary tree, whether it is satisfied is decidable.
    \textbf{TODO: cite}.
\end{proof}

@@

\begin{corollary}
    Let $L \subseteq \Sigma^*$ be an infinite language that is \kl{well-quasi-ordered} by
    the \kl{prefix relation}. Then, there exists finite numbers $k,\ell \in \Nat$ such that
    the
    \kl{maximal order type} of $L$ is $k \cdot \omega$,
    the \kl{ordinal height} of $L$ is $\omega$, and its
    \kl{ordinal width} is $\ell$.
\end{corollary}

Let us conclude by noting that it is unsurprisingly not possible to decide
whether a decidable language is \kl{well-quasi-ordered} by the \kl{prefix
relation}. This is a very easy result whose sole purpose is to contrast with
the decidability result of \cref{prefix-wqo-reg-decidable:cor}.

\begin{lemma}
    The following problem is undecidable: given a language $L$
    decided by a Turing machine, answer whether 
    $L$ is \kl{well-quasi-ordered} for the \kl{prefix relation}.
\end{lemma}
\begin{proof}
    We reduce the halting problem on the empty string $\varepsilon$.
    Let $M$ be a Turing Machine, we write the languages $L$ of finite runs
    of $M$ starting on the empty string,
    that we surround by special markers. This language is decidable,
    and 
    is
    \kl{well-quasi-ordered} if and only if it is finite
    if and only if $M$ terminates on $\varepsilon$.
\end{proof}
